Leadership across cultures and contexts: an exploratory analysis

### 2020-05-16 leadership_across_cultures_contexts.Rmd
Working on results section.

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprises `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension in the study.

@@

Subsistence strategy and group structure were also confounded. Hunter-gatherer text records often referred to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely referred to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other hand, rarely referring to residential leadership (`r agri_residential_pct`%) but often referring to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, the most frequent group type was political groups at the community level, with the exception of agriculturalists in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

We explored whether three features of our corpus of ethnographic texts -- date of publication, female gender of any author or co-author, and total pages of ethnography from each culture -- were associated with discovery of evidence for the `r length(all_study_vars)` functions, qualities, benefits, and costs dimensions (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars)). Dates of publication ranged from `r min(documents$d_publication_date)` to `r max(documents$d_publication_date)`, with a median year of `r median(documents$d_publication_date)` (see Figure S\@ref(fig:doc-years)). There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine (some authors were co-authors). Of the `r nrow(documents)` source documents, `r sum(documents$female_coauthor)` (`r signif(100*sum(documents$female_coauthor)/nrow(documents), 2)`%) had a female author or co-author. Total pages of ethnography on each culture in the PSF ranged from `r min(leader_cult[['ehraf_pages']])` to `r as.integer(max(leader_cult[['ehraf_pages']]))`, with a median of `r median(leader_cult[['ehraf_pages']])`.

The number of text records on leadership retrieved for a particular culture was weakly correlated with the total number of pages of ethnography in the eHRAF (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). A loess regression suggested that the number of text records initially increased with the total number of pages of ethnography, plateauing at about 35 records around 4500 pages (see Figure S\@ref(fig:eHRAFpages)).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary dimension as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and supplementary information for details.

We then tested if text records with a female author or co-author were more likely to discuss female leaders. Of the `r sum(tab_coauthor['FALSE',])` text records with no female author or co-author, `r tab_coauthor['FALSE','yes']` (`r 100*signif(tab_coauthor['FALSE','yes']/sum(tab_coauthor['FALSE',]),2)`%) reported female leadership. Of the `r sum(tab_coauthor['TRUE',])` text records with a female author or co-author, `r tab_coauthor['TRUE','yes']` (`r 100*signif(tab_coauthor['TRUE','yes']/sum(tab_coauthor['TRUE',]),2)`%) reported female leadership, suggesting that female ethnographers were more likely to report female leaders. After accounting for the non-independence of text records (i.e., multiple records from the same document) using a logistic mixed effects regression model, however, we found that although there was a positive effect of co-author gender on reporting female leadership (OR = `r signif(mm_coauthorOR, 2)`), it was not statistically significant (p = `r signif(mm_coauthorP, 2)`).

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing various social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders were also commonly described as serving as a representatives in inter-group interactions, managing the distribution of resources, and providing context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Notable infrequent functions included controlling group movement and immigration.

The most common qualities of leaders were having high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of cultures sampled leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained coercive authority over followers. Notable infrequent qualities of leaders included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

The frequency of leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. In these data leaders most commonly received benefits in the form of greater access to material resources (documented in `r leader_benefit_material`% of cultures) and increased social status (documented in `r leader_benefit_status`% of cultures). Mating benefits and social services benefits were also relatively frequent benefits of leadership. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in `r leader_cost_status`% of cultures) and greater access to resources (documented in `r leader_cost_material`% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems through receiving material resources (documented in `r follower_benefit_material`% of cultures) but otherwise most frequently through receiving social services (documented in `r follower_benefit_social`% of cultures). Mating and fitness benefits were relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost (documented in `r follower_cost_material`% of cultures) and in providing social services (documented in `r follower_cost_social`% of cultures). See Figure \@ref(fig:varsupportcb).

@@

Leadership can evolve when the benefits exceed the costs, and the same for followership. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader benefit/cost dimension and each follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., *Mating*, *Food*), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

@@

```{r benefitcostratio, fig.width=10, fig.cap="The odds ratios of finding evidence of a benefit vs. a cost in a text record for leaders and followers, by each dimension of benefits and costs. Odds ratios estimated using a mixed effects logistic regression model of evidence for (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. Separate models were fit for leaders and followers. X-axis is on a log scale."}
ggsave(file='Figures/plot_bc_OR.pdf', plot_bc_OR, width = 9, height = 5)
plot_bc_OR
```

@@

To investigate variation in our `r length(all_study_vars)` dimensions of leadership, we fit a multiple logistic mixed effects regression model of each dimension as a function of four factors: group context, subsistence strategy, continental region, and leader sex, with random intercepts for author nested within culture. We then compared the AIC value of each model to a model with only the fixed and random intercepts. Models of leadership dimensions whose AIC values that showed less than a two-point reduction relative to the intercept-only model were deemed to not vary by these factors, whereas those whose AIC values were less two or more were deemed to vary.

>>>>

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension in the study.

@@

Subsistence strategy and group structure were also confounded. Hunter-gatherer text records often referred to residential leadership (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and kin leadership (`r hg_kin_pct`%), but rarely referred to supracommunity leadership (`r hg_supra_pct`%). Agricultural text records, on the other hand, rarely referring to residential leadership (`r agri_residential_pct`%) but often referring to supracommunity leadership (`r agri_supra_pct`%). Across all subsistence types, political groups at the community level were the most frequent type, with the exception of agriculturalists where supracommunity political groups were the most frequent type. Leadership beyond the community level was relatively rare outside of agricultural populations and was especially rare among hunter-gatherers. See Figure \@ref(fig:grouptype).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by date of publication of the source ethnographic texts, female gender of any author or co-author, or total pages of ethnography from each culture. Dates of publication ranged from `r min(documents$d_publication_date)` to `r max(documents$d_publication_date)`, with a median year of `r median(documents$d_publication_date)` (see Figure S\@ref(fig:doc-years)). There were `r tab_author_gender['male']` male authors, `r tab_author_gender['female']` females authors, and `r tab_author_gender['unkown']` authors whose gender we could not determine (some authors were co-authors). Of the `r nrow(documents)` source documents, `r sum(documents$female_coauthor)` (`r signif(100*sum(documents$female_coauthor)/nrow(documents), 2)`%) had a female author or co-author. Total pages of ethnography on each culture in the PSF ranged from `r min(leader_cult[['ehraf_pages']])` to `r as.integer(max(leader_cult[['ehraf_pages']]))`, with a median of `r median(leader_cult[['ehraf_pages']])`, and this was weakly correlated with the number of text records on leadership retrieved for a particular culture (r = `r signif(cor_pages_tr$estimate, 2)`, p = `r signif(cor_pages_tr$p.value, 2)`). See Figure S\@ref(fig:eHRAFpages).

We fit `r length(all_study_vars)` logistic mixed effects models with each binary leadership dimension as the outcome; fixed effects for publication date, any female authors (True/False), and total pages of ethnography from each culture; and random intercepts for ethnographers nested within cultures. We then corrected for multiple comparisons by setting a false discovery rate of 5% [@benjamini1995]. After correction, none of the coefficients of female authorship or total pages of ethnography were significant, but the publication date coefficient was significant in nine models. More recent publications were more likely to report evidence for *Oratory skill*, *Group representative*, *Group determination/cohesiveness*, and *Knowledgeable/intelligent*, and less likely to report evidence for leaders imposing costs (*Punishment*, *Killer*) or receiving benefits (*Social services*, *Food*, and *Reduced risk of harm*). See Figure S\@ref(fig:pubdate) and supplementary information for details.

We then tested if text records with a female author or co-author were more likely to discuss female leaders. Of the `r sum(tab_coauthor['FALSE',])` text records with no female author or co-author, `r tab_coauthor['FALSE','yes']` (`r 100*signif(tab_coauthor['FALSE','yes']/sum(tab_coauthor['FALSE',]),2)`%) reported female leadership. Of the `r sum(tab_coauthor['TRUE',])` text records with a female author or co-author, `r tab_coauthor['TRUE','yes']` (`r 100*signif(tab_coauthor['TRUE','yes']/sum(tab_coauthor['TRUE',]),2)`%) reported female leadership, suggesting that female ethnographers were more likely to report female leaders. After accounting for the non-independence of text records (i.e., multiple records from the same document) using a logistic mixed effects regression model, however, we found that although there was a positive effect of co-author gender on reporting female leadership (OR = `r signif(mm_coauthorOR, 2)`), it was not statistically significant (p = `r signif(mm_coauthorP, 2)`). Because we found minimal evidence of bias, we do not control for any bias variables in subsequent analyses.

@@

The most common functions of leaders, documented in over 70% of cultures, were resolving conflicts, providing miscellaneous social functions, organizing various cooperative activities, and providing counsel or direction in follower's affairs. Leaders were also commonly described as serving as a representatives in inter-group interactions, managing the distribution of resources, and providing context-specific leadership such as military command and in providing ritual services. In about half of the cultures sampled here leaders punished other group members and managed economic systems. Functions notable for their rarity included controlling group movement and immigration.

The most common qualities of leaders were having high social status (documented in over 90% of cultures), being experienced or accomplished, and being described as knowledgeable or intelligent (documented in approximately 80% of cultures). Also, in about 60% of cultures sampled leaders were wealthy, nepotistic, possessed supernatural qualities, were generous, or maintained coercive authority over followers. Qualities notable for their relative infrequency included physical formidability, having many social contacts, and being feared (documented in about 35% of cultures) and charisma and fairness (documented in about 20% of cultures).

@@

Evidence for leader and follower costs and benefits were estimated similarly to the leader qualities and functions described above. Common leader benefits included greater access to material resources (documented in `r leader_benefit_material`% of cultures) and increased social status (documented in `r leader_benefit_status`% of cultures). Evidence for mating benefits and social services benefits was also relatively common. In general, the costs tracked the benefits. Leaders were often described as losing their high social status (documented in `r leader_cost_status`% of cultures) and greater access to resources (documented in `r leader_cost_material`% of cultures). See Figure \@ref(fig:varsupportcb).

Followers on the other hand most commonly benefited from leadership systems by receiving material resources (documented in `r follower_benefit_material`% of cultures) but otherwise by receiving social services (documented in `r follower_benefit_social`% of cultures). Evidence of mating and fitness benefits was relatively rare for followers. Followers were most frequently described as incurring costs in the form of material resources lost (documented in `r follower_cost_material`% of cultures) and in providing social services (documented in `r follower_cost_social`% of cultures). See Figure \@ref(fig:varsupportcb).

@@

Leadership can evolve when the benefits exceed the costs, and the same for followership. Our data did not allow us to compare benefits and costs. Instead, we compared levels of evidence for benefits vs. costs for each leader benefit/cost dimension and each follower benefit/cost dimension. Specifically, we fit a mixed effects logistic regression model of evidence (0/1) as a function of the leader dimension (e.g., *Mating*, *Food*), a binary term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. We fit separate models for leaders and followers. These two models allowed us to estimate the relative evidence for, e.g., leader mating benefits vs. leader mating costs, and follower mating benefits vs. follower mating costs, etc. For leaders, the odds ratio that a text record would provide evidence for a benefit vs. a cost (averaging across all benefit/cost dimensions), was $OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$, and for followers, was $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$. Hence for both leaders and followers, there was more evidence of benefits than costs.

@@

```{r benefitcostratio, fig.width=10, fig.cap="The odds ratios of finding evidence of a benefit vs. a cost in a text record for leaders and followers, by each dimension of benefits and costs. Odds ratios estimated using a mixed effects logistic regression model of evidence (0/1) as a function of each dimension (e.g., Mating, Food), a term indicating if the evidence was for a benefit or cost, and their interaction, with random intercepts for author nested within culture. Separate models were fit for leaders and followers. X-axis is on a log scale."}
ggsave(file='Figures/plot_bc_OR.pdf', plot_bc_OR, width = 9, height = 5)
plot_bc_OR
```

@@

We investigated four possibles sources of variation in evidence for our `r length(all_study_vars)` dimensions of leadership: group context, subsistence strategy, continental region, and leader sex. Specifically, we fit separate multiple logistic mixed effects regression models of each dimension as a function of these four factors, with random intercepts for author nested within culture. We then compared the AIC value of each model to a model with only the fixed and random intercepts. Models of leadership dimensions whose AIC values that showed less than a two-point reduction relative to the intercept-only model were deemed to not vary by these factors, whereas those whose AIC values were less two or more were deemed to vary [@burnham2002].
